\section{Our technique}
\label{sec-our-technique}

For the purpose of this work, we define a \emph{function call} to be
the code that accomplishes the following tasks:

\begin{enumerate}
\item It accesses the arguments to be passed to the callee from the
  places they have been stored after computation, and puts the
  arguments in the places that the callee expects them.
\item It accesses the function object associated with the name at the
  call site and stores it in some temporary location.
\item From the function object, it accesses the static environment to
  be passed to the code of the callee.
\item Also from the function object, it accesses the \emph{entry
  point} of the function, i.e., typically the address of the first
  instruction of the code of the callee.
\item It transfers control to the entry point, using an instruction
  that saves the return address for use by the callee to return to the
  caller.
\item Upon callee return, it accesses the return values from the places
  they have been stored by the callee, and puts those values in the
  places where the caller requires them for further computation.
\end{enumerate}

In a typical implementation, a function call is generated when the
code of the caller is compiled, and it then never changes.  For this
permanent code to work, a particular \emph{call protocol} must be
observed, and that protocol must be independent of the callee, as the
callee may change after the caller has been compiled.

With our suggested method, the function call is created by the
callee.  The code emitted by the caller for a function call consists
of a single unconditional \emph{jump} instruction.  The target address
in that instruction is altered by the callee according to its
structure.  The code for the function call is contained in an object
that we call a \emph{snippet}.  The callee allocates an appropriate
snippet in the global heap as describe in \refSec{sec-sicl-features},
at some available location, and the unconditional jump instruction of
the caller is modified so that it performs a jump to the first
instruction of the snippet.  When the callee changes in some way, a
new snippet is allocated and the jump instruction is altered to refer
to the position of the new snippet.  The old snippet is then subject
to garbage collection like any other object.

In order for the callee to be able to adapt the snippet to its
requirements, the caller, when loaded into the executing image, must
provide information about its call sites to the system.  Each call
site contains information such as:

\begin{itemize}
\item The number of arguments.
\item The type of each argument.  If the type is not known, it is
  indicated as \texttt{t}.  When an argument is a literal object, its
  type is indicated as \texttt{(eql ...)}.
\item For each argument, whether the argument is boxed or unboxed.
\item For each argument, its location.  The location can be a register
  or a stack position in the form of an offset from a frame pointer.
\item The number of required return values, or an indication that all
  return values are required, no matter the number.
\item In case of a fixed number of return values, for each such value,
  whether it should be boxed or unboxed.
\item Also, in case of a fixed number of return values, for each such
  value, the location where the caller expects the value.
\item Indication as to whether the call is a \emph{tail call}, in
  which case the snippet should deallocate the frame before
  returning.
\end{itemize}

A callee can take advantage of this information to customize the
call.  The default action is to generate a snippet that implements the
full function-call protocol, without taking into account information
about the types of the arguments.

When a modification is made to a callee that alters its semantics,
care must be taken so as to respect the overall semantics of all
callers.  In particular, a callee can be removed using
\texttt{fmakunbound} or entirely replaced using \texttt{(setf
  fdefinition)}.  In that case, the following steps are taken in
order:

\begin{enumerate}
\item First, a new snippet is allocated for each caller, and the
  unconditional jump instruction is modified to refer to the new
  snippet.  This snippet contains code for the full function-call
  protocol in that it puts the arguments in the places defined by that
  protocol, and in that it accesses the callee using an indirection
  through the function cell.
\item Next, the callee is atomically replaced by a new function, or
  entirely removed by a single modification to the contents of the
  function cell.
\item The new function is attached to the list of call sites, and,
  depending on the nature of the new function, new snippets can then
  be allocated in order to improve performance of calls to the new
  function.
\end{enumerate}

%%  LocalWords:  callee
