\chapter{Condition system}

This module is located in the \texttt{Code/Conditions} directory.  It
it meant to contain the complete \commonlisp{} condition system.  The module
is fairly complete.  There are three main items in this module

\begin{itemize}
\item A complete portable implementation of the \commonlisp{} condition
  system signaling mechanism.
\item Definitions of standard \commonlisp{} conditions.
\item Definitions of additional \sysname{} conditions.
\end{itemize}

As it turns out, the \commonlisp{} condition signaling mechanism can be
implemented in a completely portable way, using only special variables
as the underlying mechanism.  In addition, this implementation method
is likely to exhibit acceptable performance, because condition
signaling is assumed to be fairly infrequent anyway.  

The module assumes the existence of the \texttt{define-condition}
macro.  The full hierarchy of \commonlisp{} conditions is implemented, and
there is a file containing condition reporters for those conditions
when the English language is used.

The additional conditions are used by various \sysname{} modules.
At the moment, this module does not contain all additional conditions
used by the system.  We are still undecided concerning whether all
additional conditions should be concentrated in one place, or on the
contrary, they should be distributed with the particular module that
uses them.  The first solution favors sharing of conditions that are
usable by several modules, and the second solution lets the user of a
module install only the conditions used by that module, without having
to install a large number of conditions that will never be used.
