\chapter{Medium-level intermediate representation (MIR)}
\label{chap-mir}

The compiler translates an abstract syntax tree into a \emph{graph} of
\emph{instructions} in a language named \textbf{MIR}.  It is a general
graph because an instruction can have zero, one, or more
\emph{successors}, and zero, one, or more \emph{predecessors}.  The
graph can also contain \emph{cycles}.  

An instruction can also have zero, one, or several \emph{inputs} and
zero, one, or more \emph{outputs}.  An input may be an
\emph{immediate} value, a \emph{lexical variable}, or an \emph{entry
  in the linkage vector}.  An output may only be a lexical variable.

Executing a MIR instruction consists of generating the outputs as a
function of the inputs, and also of choosing a successor based on the
inputs.

An execution of a MIR program consists of executing a sequence of
instructions where an instruction in the sequence is the successor
chosen during the execution of the preceding instruction in the
sequence.

\section{Definition of MIR instructions}

\subsection{Instruction \texttt{NOP}}
\label{mir-instruction-NOP}

This instruction has a single successor.  It has no inputs and no
outputs.

Executing this instruction has no effect. 

\subsection{Instruction \texttt{assignment}}
\label{mir-instruction-assignment}

This instruction has a single successor.  It has a single input and a
single output. 

The execution of this instruction results in the input being copied to
the output without any modification. 

\subsection{Instruction \texttt{test}}
\label{mir-instruction-test}

This instruction has two successors.  It has a single input. 

The first successor is chosen if the input is \emph{true} and the
second input is chosen if the input is \emph{false}.%
\fixme{Check that this statement is true.}

\subsection{Instruction \texttt{funcall}}
\label{mir-instruction-funcall}

This instruction has a single successor.  It has a single input and no
outputs.

The input contains the function to be called.  Arguments must already
have been communicated by the use of the \texttt{put-arguments}
instruction. \seesec{mir-instruction-put-arguments}

\subsection{Instruction \texttt{get-arguments}}
\label{mir-instruction-get-arguments}

This instruction has a single successor.%
\fixme{Say more.}

\subsection{Instruction \texttt{get-values}}
\label{mir-instruction-get-values}

This instruction has a single successor.  It has no inputs and it can
have any number of outputs.

The effect of this instruction is to assign the values of a preceding
function call to a set of lexical variables. 

\subsection{Instruction \texttt{put-arguments}}
\label{mir-instruction-put-arguments}

This instruction has a single successor.  It can have any number of
inputs, and it has not outputs. 

The effect of this instruction is to communicate the value of the
inputs as arguments to the next \texttt{funcall}
instruction. \seesec{mir-instruction-funcall}

\subsection{Instruction \texttt{put-values}}
\label{mir-instruction-put-values}

This instruction has a single successor.  It can have any number of
inputs, and it has no outputs. 

The effect of this instruction is to communicate the value of the
inputs as values to be returned when this function finishes its
execution.

\subsection{Instruction \texttt{enter}}
\label{mir-instruction-enter}

\subsection{Instruction \texttt{leave-values}}
\label{mir-instruction-leave-values}

\subsection{Instruction \texttt{return}}
\label{mir-instruction-return}

This instruction has no successors, no inputs and no outputs.

It terminates execution of the current function and returns to the
caller. 

\subsection{Instruction \texttt{enclose}}
\label{mir-instruction-enclose}

This instruction has a single successor.  It has a single output. 
It does have an input, but it is a special one because it is not a
value, but instead the root of an instruction graph.  

The instruction takes the instruction graph and creates a
\emph{closure}.  The closure contains the current lexical runtime
environment and the code resulting from the input instruction graph.

\subsection{Instruction \texttt{memref}}
\label{mir-instruction-memref}

This instruction has a single successor.  It has one input and one
output.  

The input is a raw memory address, and the output is the contents of
memory at that address. 

\subsection{Instruction \texttt{memset}}
\label{mir-instruction-memset}

This instruction has a single successor.  It has two inputs and no
outputs. 

The first output is a memory address and the second input is a value
to be stored in memory at that address. 

\subsection{Instruction \texttt{u+}}
\label{mir-instruction-u+}

This instruction has one or two successors.  It has two inputs and one
output.

The instruction computes the unsigned addition of its two inputs.  The
output is the sum modulo the $2^w$ where $w$ is the width of a word in
bits.  If the instruction has two successors, then the first one is
taken if there is no \emph{overflow} and the second one is taken if
there is.  

\subsection{Instruction \texttt{u-}}
\label{mir-instruction-u-}

This instruction has one or two successors.  It has two inputs and one
output.

The instruction computes the unsigned subtraction of its two inputs.
The output is the difference modulo the $2^w$ where $w$ is the width
of a word in bits.  If the instruction has two successors, then the
first one is taken if there is no \emph{overflow} and the second one
is taken if there is.

\subsection{Instruction \texttt{s+}}
\label{mir-instruction-s+}

This instruction has one or two successors.  It has two inputs and one
output.

The instruction computes the signed addition of its two inputs.  The
output is the sum modulo the $2^w$ where $w$ is the width of a word in
bits.  If the instruction has two successors, then the first one is
taken if there is no \emph{overflow} and the second one is taken if
there is.  

\subsection{Instruction \texttt{s-}}
\label{mir-instruction-s-}

This instruction has one or two successors.  It has two inputs and one
output.

The instruction computes the signed subtraction of its two inputs.
The output is the difference modulo the $2^w$ where $w$ is the width
of a word in bits.  If the instruction has two successors, then the
first one is taken if there is no \emph{overflow} and the second one
is taken if there is.

\subsection{Instruction \texttt{neg}}
\label{mir-instruction-neg}

This instruction has one or two successors.  It has one input and one
output.

The instruction computes the negation of its input.  The output is the
negation modulo the $2^w$ where $w$ is the width of a word in bits.
If the instruction has two successors, then the first one is taken if
there is no \emph{overflow} and the second one is taken if there is.

\subsection{Instruction \texttt{\&}}
\label{mir-instruction-and}

This instruction has a single successor.  It has two inputs and one
output. 

This instruction computes the bitwise logical \emph{and} between the
two inputs. 

\subsection{Instruction \texttt{ior}}
\label{mir-instruction-ior}

This instruction has a single successor.  It has two inputs and one
output. 

This instruction computes the bitwise logical \emph{inclusive or}
between the two inputs.

\subsection{Instruction \texttt{xor}}
\label{mir-instruction-xor}

This instruction has a single successor.  It has two inputs and one
output. 

This instruction computes the bitwise logical \emph{exclusive or}
between the two inputs.

\subsection{Instruction \texttt{\~}}
\label{mir-instruction-not}

\subsection{Instruction \texttt{s<}}
\label{mir-instruction-s-less}

\subsection{Instruction \texttt{s<=}}
\label{mir-instruction-s-less-or-equal}

\subsection{Instruction \texttt{u<}}
\label{mir-instruction-u-less}

\subsection{Instruction \texttt{u<=}}
\label{mir-instruction-u-less-or-equal}

\subsection{Instruction \texttt{catch}}
\label{mir-instruction-catch}

\subsection{Instruction \texttt{unwind}}
\label{mir-instruction-unwind}
