\chapter{High-level functions on lists}

This module is meant to be a complete implementation of portable
functions and macros in the Conses dictionary (chapter 14 in the
\hs{}), except for the low-level functions such as \texttt{cons},
\texttt{car}, \texttt{cdr}, \texttt{rplaca}, and \texttt{rplacd} which
can not be implemented portably.  For its implementation, it uses the
\texttt{loop} macro.  If any other functionality is required, it will
supply special implementations of such functionality, so as to avoid
dependencies on other modules.

We obtain high performance by identifying important special
cases such as the use of \texttt{:test} function \texttt{eq}, or
\texttt{equal}, or the use of a \texttt{:key} of \texttt{identity}.

We supply compiler macros so as to avoid runtime dispatch whenever a
special-case function can be determined by only looking at the call
site.  This ensures high performance for short lists, where argument
parsing would otherwise represent a significant fraction of the cost
of the call.

We are considering whether it might be worthwhile to supply a
macroexpanded version of this module so that no existing
implementation of the \texttt{loop} macro would be required. 

This module is fairly complete, and it includes macros such as
\texttt{push} and \texttt{pop} as well as compiler macros for
functions that take keyword arguments such as the mapping functions. 

The module also includes definitions of specific conditions that are
used by this module, together with condition reporters in English for
those conditions.  It also includes English-language documentation
strings for some of the functions. 
