\section{Conclusions and future work}
\label{sec-conclusions}

We have defined a technique that solves a problem encountered in
current \commonlisp{} implementations when a \texttt{defgeneric} form
is followed by a \texttt{defmethod} form in the same compilation
unit.  When the \texttt{defgeneric} form mentions a method class other
than \texttt{standard-method}, and the compilation unit is processed
in a fresh compilation environment, current implementations do not
propagate the information about the method class to the macro expander
for \texttt{defmethod}, resulting in \texttt{make-method-lambda} being
called with a \texttt{method} argument of the wrong class.

Our solution requires the compiler of the \commonlisp{} implementation
to store a small amount of additional information about the
generic-function class and the method class when the
\texttt{defgeneric} form is encountered, and requires the macro
expander for \texttt{defmethod} to retrieve this information by
querying the compilation environment.

Contrary to the proposal by \cnh{}, our suggested solution does not
introduce any incompatibilities that would render some existing code
obsolete.  Furthermore, our solution does not have the potential
performance problem of the proposal by \cnh{}, i.e. the additional
cost of processing keyword arguments to method functions.

Future work includes adding the functions defined in
\refApp{app-protocol} to the \sicl{}%
\footnote{See https://github.com/robert-strandh/SICL} protocol for
first-class global environments.

The \cleavir{} compiler framework which is part of \sicl{} defines a
modern version of the protocol for environment query defined in the
second edition of Guy Steele's book \cite{Steele:1990:CLL:95411}.  We
plan to extend this protocol to include information about the name of
the generic-function class and of the method class given (explicitly
or implicitly) in a \texttt{defgeneric} form previously encountered in
the current compilation environment.  Since our existing protocol
returns standard objects, no modifications to the existing \cleavir{}
code will be required as a result of this extension.  The extension
will allow us to define the macro \texttt{defmethod} in \sicl{} to
query the environment, and to invoke \texttt{make-method-lambda} with
appropriate arguments.
