\chapter{Introduction}
\pagenumbering{arabic}%

\sysname{} is an implementation-independent framework for creating
\cl{} compilers. 

It is implementation-independent in that it provides:

\begin{itemize}
\item general features that every implementation needs,
\item features that implementations can optionally choose to take
  advantage of, 
\item alternative features that are appropriate for some
  implementations and not for others,
\item mechanisms for allowing implementation-specific features that
  integrate seamlessly into the general framework.
\end{itemize}

To use the framework, an implementation would translate source code
into an \emph{abstract syntax tree}, or AST.
\seechap{chap-abstract-syntax-tree}
\sysname{} then translates the AST into an intermediate representation
called MIR.  \seechap{chap-mir}
This translation can be customized by the implementation.
\sysname{} then provides a number of transformations on this
representation such as:

\begin{itemize}
\item translation into \emph{static single assignment} form,
\item type inference,
\item standard optimizations such as value numbering, common
  subexpression elimination, redundancy elimination, etc.
\end{itemize}

Once these implementation-independent and backend-independent
transformations have been accomplished, the MIR notation is gradually
transformed into an notation that is specific both to the
implementation and to the backend in that:

\begin{itemize}
\item it introduces backend-specific features such as registers and
  stack frames,
\item it exposes implementation-specific choices such as argument
  passing.
\end{itemize}

Finally, the low-level code is translated into machine code.

At each stage, an implementation can customize the process by
introducing new classes and methods. 
