\section{Type inference method}

Because \commonlisp{} types are arbitrary sets, perfect inference of
\commonlisp{} types would be equivalent to the halting
problem\footnote{Most obviously, if the type of the input to a
  function is a singleton set.}; additionally, some type information
is more useful than others in an optimizing compiler. Therefore, a
small finite lattice is used for type inference.

Elements of the lattice are called \textit{type descriptors}, or just
``descriptors''. Every Common Lisp type can be conservatively
approximated by a descriptor. That is, the type will not contain any
elements not also contained in the descriptor. There is a descriptor
for the top type, bottom type, and any type that is useful for
optimization.

Descriptors are collected into \textit{bags}, which are associative
maps from variables to descriptors. Bags are equipped with equality
and union operations, which are both only defined for bags with the
same set of variables, and consist of the equivalent operation on
descriptors mapped over the variables.

Type inference begins by associating each arc with a bag. Initially,
every descriptor in every bag is the top type descriptor. As an
optimization, only those variables which are \textit{live} (used as
inputs in some instruction later in control flow) during the arc are
included in each bag.

This map of arcs to bags represents all useful type information, and
an altered map is the ultimate result of the type inference
procedure. Type inference consists of successively shrinking the
descriptors of variables in the bags from the top descriptor.

To proceed with type inference, all instructions with predecessors are
iterated over in arbitrary order. For each instruction, a specialized
\textit{transfer function} is executed. The transfer function receives
as arguments the instruction and a bag. The bag is the union of bags
in arcs directly preceding the instruction. The transfer function then
computes a bag for each succeeding arc of the instruction based on
these arguments.

If the bags so computed are distinct from the bags already associated
with those arcs, new information is available. The succeeding
instructions are then marked to be iterated over again. This continues
until no instructions are left to iterate over, and at this point the
current association of arcs with bags is the result of type inference.

For example, a \commonlisp{} programmer may annotate their program
with type information by using the \texttt{the} special operator:
\texttt{(the fixnum foo)} indicates that whenever this form is
evaluated, the variable \texttt{foo} is of type
\texttt{fixnum}. \texttt{the} is represented in HIR by a
\texttt{the-instruction}, which has one predecessor, one successor,
and one input, the variable \texttt{foo}. During type inference, the
transfer function for the \texttt{the-instruction} returns the
information that in the successor arc, \texttt{foo} must fit the type
descriptor most closely approximating the \commonlisp{} type
\texttt{fixnum}, and transfer functions for instructions farther along
in the control flow can propagate this information.
